
Red Wine Quality Exploration by Mahmoud Esmail
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# importing the necessary packages to our workspace for future work
library(ggplot2)
library(dplyr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

RedWine <- read.csv('wineQualityReds.csv')
```

> **Dataset Introduction**: We are going to find correlations between the 11 features in RedWine dataset and after analyzing it. we will find factors that affects wine quality

```{r echo=FALSE}
#Statistical summary of the data set
summary(RedWine)
```
```{r echo=FALSE}
#structures of objects in the dataset
str(RedWine)
```
**Summary of the dataset**:

* we have 1599 object and 12 variable at which 
11 of them are features that describes quality of wine
* it turns out that the max quality of wine reaches
8.0 and no quality is higher than that, while the average quality is 5.635


# Univariate Plots Section

* in this section I'm going to examine the distribution- 
* of each variable in this dataset by plots to check
whether it's normally distributed or not

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots1}
#plotting every 12 feature in the dataset to find 
#their distributions and any interesting trends

ggplot(aes(x=fixed.acidity), 
       data=RedWine)+ geom_histogram()+
  ggtitle('Fixed Acidity Distribution')

```

**Observations**

* fixed acidity distribution is long tailed we need to fix it

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}
 ggplot(aes(x=volatile.acidity), data=RedWine)+ 
  geom_histogram() +
  ggtitle('Volatile Acidity Distribution')
```

**Observations**

* volatile acidity distribution is long tailed we need to fix it

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}
 ggplot(aes(x=citric.acid), data=RedWine)+ 
  geom_histogram() + 
  ggtitle('Citric Acid Distribution')
```

**Observations**

* citric acid distribution is positively skewed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}
ggplot(aes(x=residual.sugar), data=RedWine)+ 
  geom_histogram() +
  ggtitle('Residual Sugar Distribution')
```

**Observations**

* residual sugar is not normally distributed, it's long tailed,
<br />
  starting from non-zero position and there's 
  <br /> a few outliers at values
  after 5 on the scale

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}
 ggplot(aes(x=chlorides), data=RedWine)+
  geom_histogram() +
  ggtitle('Chlorides Distribution')
```

**Observations**

* chlorides is long tailed disrtibution,
  there's a few outliers from 0.4 to 0.6


```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots6}
 ggplot(aes(x=free.sulfur.dioxide), data=RedWine)+
  geom_histogram() + 
  ggtitle('Free Sulfur Dioxide Distribution')
```

**Observations**

* free sulphur dioxide distribution is long tailed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7}
 ggplot(aes(x=total.sulfur.dioxide),
        data=RedWine)+
  geom_histogram() + 
  ggtitle('Total Sulfur Dioxide Distribution')
```

**Observations**

* total sulphure dioxide is long tailed distribution
it need to be added on   <br /> 
log10 scale to see if it
could be normally distributed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots8}
 ggplot(aes(x=density), data=RedWine)+ 
  geom_histogram() + 
  ggtitle('Density Distribution')
```

**Observations**

* when we take a look density plot 
  we realize that it's normally distributed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots9}
 ggplot(aes(x=pH), data=RedWine)+
  geom_histogram() + 
  ggtitle('pH Distribution')
```

**Observations**

* when we take a look pH plot 
  we realize that it's normally distributed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots10}
 ggplot(aes(x=sulphates), data=RedWine)+
  geom_histogram() + 
  ggtitle('Sulphates Distribution')
```

**Observations**

* sulphates distribution is long-tailed

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots11}
 ggplot(aes(x=alcohol), data=RedWine)+ 
  geom_histogram() +
  ggtitle('Alcohol Distribution')
```

**Observations**

* there's a few outliers before 9 and after the 14
* the histogram could be divided into two plots before 
  and after 11
  on the scale, after dividing it   <br /> 
  we see that the two plots are not
  normally distributed



```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots12}
 ggplot(aes(x=quality), data=RedWine)+
  geom_histogram() +
  ggtitle('Quality Distribution')
```

**Observations**

* most of wine qualities are between 5 and 6 on average, 
  where lowest quality   <br />
  is nearly about 3 and highest quality is nearly 8 



  

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_13}
p1 <- ggplot(aes(x=fixed.acidity), data=RedWine)+ 
  geom_histogram() + 
  ggtitle('Fixed Acidity Distribution') + 
  scale_x_log10(breaks=seq(0,16))

p2 <- ggplot(aes(x=volatile.acidity), data=RedWine)+ 
  geom_histogram() +
  ggtitle('Volatile Acidity Distribution')+
  scale_x_log10(breaks=seq(0,1.6,0.2))

p3 <- ggplot(aes(x=total.sulfur.dioxide), data=RedWine)+ 
  geom_histogram() + 
  ggtitle('Total Sulfur Dioxide Distribution') + scale_x_log10(breaks=seq(0,300,100))

p4 <- ggplot(aes(x=sulphates), data=RedWine)+
  geom_histogram() +
  ggtitle('Sulphates Distribution') +
  scale_x_log10(breaks=seq(0,2,0.5))

grid.arrange(p1,p2,p3,p4, ncol=2)

```

**Observations**

* Now the plots became more normally distributed, 
but still there's a small outliers


# Univariate Analysis


### What is the structure of your dataset?

* the dataset contains 12 main variables or objects 
each of which describes
  the quality of red wine

* Actually, the 12 variables 11 of them are
experimental variables or 

    in other words they're measurements to describe
    the attributes of wine 

     while the one left variable which is "quality" 
     is a variable which is  

     rated from 0 to 10 by wine experts
  
* the worst wine is rated nearly 3 while the best is about 8


### What is/are the main feature(s) of interest in your dataset?

the main feature in our dataset is Quality of the red wine

### What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?

most of variables(features) in the dataset could
support the investigation
but the most powerful features 

that may support the investigation are pH, Alcohol 
and residual sugar

### Did you create any new variables from existing variables in the dataset?
No

### Of the features you investigated, were there any unusual distributions? 
Did you perform any operations on the data to tidy, adjust, or change the form
of the data? If so, why did you do this?

* I performed adjustments on the distributions 
that doesn't seem to be normal  

(fixed.acidity, volatile.acidity, 
total sulphur dioxide, sulphates),  
I did so to see if they could be normally
distributed to find the mean value  

of the attribute and whether or not it has huge 
impact on the quality of red wine

* the unusual distribution appears on quality distribution 
  because there's a few values in 3, 4, 8 while   <br />
  the most
  of values ranges 
  
  between 5 and 6 which means there's a roam for unstability though.
  

# Bivariate Plots Section

**In this Plotting Section**:
I'm going to plot every single attribute against quality to see if there's a correlation between the variables 

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots1}
#plotting fixed acidity against quality
# to see if there's any correlations

ggplot(aes(x=fixed.acidity, y=quality), data = RedWine) + geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Fixed Acidity vs. Quality')
```

**Observation**:

* There's no correlation between fixed acidity and quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots2}

#finding correlations between volatile acidity and quality

ggplot(aes(x=volatile.acidity, y=quality), data = RedWine) + geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Volatile Acidity vs. Quality')
```

**Observation**:

* At some regions there's no correlation between
volatile acidity and quality 
  and at another points   <br />
  there's a weak correlation
  at which the higher 
  volatile acidity the lower quality


```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots3}

#residual sugar against quality

ggplot(aes(x=residual.sugar, y=quality), data = RedWine) + geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2)+
  ggtitle('Residual Sugar vs. Quality')
```

**Observation**:

* There's a correlation between amount of residual sugar
and quality of red wine   <br />
  but still not clear enough so it's weak correlation

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots4}

#chlorides against quality plot

ggplot(aes(x=chlorides, y=quality), data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Chlorides vs. Quality')

```

**Observation**:

* The lower the chlorides the higher the quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots5}
#citric acid against quality plot

ggplot(aes(x=citric.acid, y=quality), data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Citric Acid vs. Quality')

```

**Observation**:

* There's no clear correlation between citric acid and quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots6}

#free sulphur dioxide against quality plot

ggplot(aes(x=free.sulfur.dioxide, y=quality), 
       data = RedWine) + geom_jitter(alpha=0.5)+
  geom_smooth(alpha=0.2)+
  ggtitle('Free Sulphur Dioxide vs. Quality')
```

**Observation**:

* There's no correlation between free sulphur dioxide and quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots7}
#total sulphur dioxide against quality plot
ggplot(aes(x=total.sulfur.dioxide, y=quality),
       data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) + 
  ggtitle('Total Sulphur Dioxide vs. Quality')
```

**Observation**

* there's no clear correlation between total sulphur dioxide and quality


```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots8}
#density against quality plot

ggplot(aes(x=density, y=quality), data = RedWine) + 
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Density vs. Quality')
```

**Observation**:

* There's no correlation between density and quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots9}
#pH against quality plot
ggplot(aes(x=pH, y=quality), data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2)+
  ggtitle('pH vs. Quality')
```

**Observation**:

* There's no strong correlation between pH and quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots10}

#sulphates against quality plot

ggplot(aes(x=sulphates, y=quality), data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Sulphates vs. Quality')

```

**Observation**:

* There's no strong correlation between sulphates and quality 
  at some points below 0.7   <br />
  the higher sulphates means higher quality 
  after that point,   <br />
  I see the higher sulphates means lower quality

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots11}

#alcohol against quality plot

ggplot(aes(x=alcohol, y=quality), data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2) +
  ggtitle('Alcohol vs. Quality')+
    xlab(expression(Alcohol~('% by volume'))) +
  ylab(expression(Quality~('From 0 to 10')))
```

**Observation**:

* at some points it appears there's a small correlation which
  implies that the 
  the higher the quality   <br />
  the higher the alcohol quantity
  
  
  
##Also I'm Going to plot some relations between input variables to see 
Correlations

```{r ,echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots12}

#some relations between input variables
#and how they affect on each other

qp1<- ggplot(aes(x=fixed.acidity, y=citric.acid),
             data = RedWine) + geom_jitter(alpha=0.5)+
  geom_smooth(alpha=0.2) + 
  ggtitle('Fixed Acidity vs. Citric Acid')+
  xlab(expression(Fixed~Acidity~(g/dm^{3}))) +
  ylab(expression(Citric~Acid~(g/dm^{3})))

qp2<- ggplot(aes(x=volatile.acidity, y=citric.acid)
             , data = RedWine) +
  geom_jitter(alpha=0.5)+ geom_smooth(alpha=0.2)+
  ggtitle('Volatile Acidity vs. Citric Acid')+
  xlab(expression(Volatile~Acidity~(g/dm^{3}))) +
  ylab(expression(Citric~Acid~(g/dm^{3})))

qp3<- ggplot(aes(x=total.sulfur.dioxide, y=free.sulfur.dioxide), 
             data = RedWine) +
  geom_jitter(alpha=0.5)+ 
  geom_smooth(alpha=0.2) + 
  ggtitle('Total sulphur dioxide vs. Free sulphur dioxides')+
  xlab(expression(Free~Sulfur~Dioxide~(mg/dm^{3}))) +
  ylab(expression(Total~Sulfur~Dioxide~(mg/dm^{3})))

qp4<- ggplot(aes(x=density, y=alcohol), data = RedWine) + geom_jitter(alpha=0.5)+
  geom_smooth(alpha=0.2) +
  ggtitle('Density vs. Alcohol')+
  xlab(expression(Density~(g/cm^{3}))) +
  ylab(expression(Alcohol~(g/dm^{3})))

#adding plots in  a grid
grid.arrange(qp1,qp2,qp3,qp4, ncol=2)
```

**Observations**

* total sulphur and free sulphur dioxide has a strong correlation
* fixed acidity and citric acid has strong correlation


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset?

* there's a small positive correlation between alcohol content and quality

* also there's a correlation between chlorides and quality,  <br />
  but negative 
  this time the lower the chlorides the higher the quality
  
* there's a small correlation between amount of residual sugar and quality
  though, it's a weak correlation
  
* volatile acidity at the first few points have positive correlations
  and negative correlations at the rest of points

* there's no other clear evidences or correlations between other attributes      which affects the quality

### Did you observe any interesting relationships between the other features 
(not the main feature(s) of interest)?

* Fixed acidity and citric acid are correlated in a positive way
* total sulphur dioxides are strongly correlated with free sulphur dioxides


### What was the strongest relationship you found?

* the strongest relationship is the relation between alcohol and quality

* another strong relation but negatively is relation between chlorides 
  and quality

# Multivariate Plots Section

* since the alcohol and chlorides seems to be the most correlatable attributes
  with quality, I will plot them against quality
  
* also residual sugar is correlatable is someway with quality so i will plot it
  with chlorides against quality
  
* from the plot, volatile acidity and citric acid seems to be uncorrelatble as   well   <br />
  so i will plot them to see if there's any interesting trends

```{r echo=FALSE, warning=FALSE, message=FALSE, Multivariate_Plots1}

# plotting some input variables against quality
# to see any interesting trends
ggplot(aes(x=alcohol, y=chlorides, color=as.factor(quality)),
       data = RedWine) + geom_point() + 
  ggtitle('Alcohol vs. Chlorides vs. Quality')+
  
  scale_color_brewer(palette = "YlGnBu",
                     direction = -1, name = "Quality") +
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))+
  xlab(expression(Alcohol~(g/dm^{3}))) +
  ylab(expression(Chlorides~(g/dm^{3})))
```

```{r echo=FALSE, warning=FALSE, message=FALSE, Multivariate_Plots2}

# plotting residual sugar and chlorides against quality
# to see any interesting trends

ggplot(aes(x=residual.sugar, y=chlorides, color=as.factor(quality)),
       data = RedWine) + geom_point() +
  ggtitle('Residual Sugar vs. Chlorides vs. Quality')+
  scale_color_brewer(palette = "YlOrBr",
                     direction = -1, name = "Quality") +
  
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey")) +
  
  xlab(expression(Residual~Sugar~(g/dm^{3}))) +
  ylab(expression(Chlorides~(g/dm^{3})))
```

```{r echo=FALSE, warning=FALSE, message=FALSE, Multivariate_Plots3}

# plotting volatile acidity and citric acid against quality
# to see any interesting trends

ggplot(aes(x=volatile.acidity, y=citric.acid, color=as.factor(quality)),
       data = RedWine) + 
  geom_point() +
  ggtitle('Volatile Acidity vs. Citric acid vs. Quality')+
  scale_color_brewer(palette = "RdYlBu", 
                     direction = -1, name = "Quality") +
  
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))+
  xlab(expression(Volatile~Acidity~(g/dm^{3}))) +
  ylab(expression(Citric~Acid~(g/dm^{3})))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest?

* when we putted down alcohol vs chlorides vs quality in a plot it looks
  like it strenghten the insights and it makes it   <br />
  comfortable to say that
  the higher alcohol with  smaller amount of chlorides   <br />
  makes the best quality rating
  
* there's no strong correlations of residual sugar with chlorides
  there's some points when   <br />
  the amount sugar increase above the limit   <br />
  and amount of chloride is few the quality decreases

* when the volatile acidity is low with a high citric acid
  the quality becomes high, and vice versa.

### Were there any interesting or surprising interactions between features?

* the surprise is that residual sugar when increases the quality becomes low

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_One}

ggplot(aes(x=alcohol, y=quality), data = RedWine)+
  geom_jitter(alpha=0.2)+
  geom_smooth(alpha=0.6) + 
  ggtitle('Alcohol vs. Quality')+
  xlab(expression(Alcohol~('% by volume'))) +
  ylab(expression(Quality~('Score between 0 and 10')))
```

### Description One

* alcohol affects the quality rating of the redwine, it's not the only factor
  that affects but   <br />
  it's the most powerful attribute that affects the final  rating


### Plot Two

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Two}
ggplot(aes(x=citric.acid, y=quality), data = RedWine) + 
  geom_jitter(alpha=0.2)+
  geom_smooth(alpha=0.8) + 
  ggtitle('Citric Acid vs. Quality')+
  xlab(expression(Citric~Acid~(g/dm^{3}))) +
  ylab(expression(Quality~('From 0 to 10')))
```

### Description Two

* citric acid seems to be a little bit 
  effective on the quality even if
  the plot doesn't <br />
  give a strong a correlation   <br />
  but one thing i noticed is that smooth wave that goes up at 0.25
  this inferes that the quality <br />
  might increase by increasing   <br />
  the amount of citric acid but it not might 
  be the only factor   <br />
  of increasing quality
  as there's another factors
  
  
### Plot Three

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Three}
ggplot(aes(x=residual.sugar, y=chlorides, color=as.factor(quality)),
       data = RedWine) + geom_point() +
  ggtitle('Residual Sugar vs. Chlorides vs. Quality')+
  scale_color_brewer(palette = "YlOrBr",
                     direction = -1, name = "Quality") +
  
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey")) +
  
  xlab(expression(Residual~Sugar~(g/dm^{3}))) +
  ylab(expression(Chlorides~(g/dm^{3})))
```

### Description Three

* in the univariate plot, When we plotted the residual.sugar against 
quality
  it does not   <br />
  gave a strong correlation that the higher amount of sugar
  implies
  higher quality of wine
* But, After adding another factor   <br />
to our graph which is chlorides 
it turns out
  that we augmented the investigation a little bit,   <br />
  we found that residual sugar when increases 
  above the limit without sufficient
  chlorides amount the wine quality decreases 
  

------

# Reflection

* I was surprised by the overall industry of red wine
  I didn't expect that it depends on many many attributes   <br />
  and combination
  of attributes and how different mixtures of inputs can affect the output
  quality of the wine   <br />
  maybe one attribute when added to the wine could have a drawback
  but when it added with the appropriate quantity   <br />
  and with another elements
  it will make amazing tastes and quality ratings
  I saw it in from the plots that many variables   <br />
  are not correlatable
  but when they are added with another variables the became extremly
  correlatble   <br />
  also, the citric acid and alcohol are mainly the most
  correlatble variables
  with wine quality

* I ran into difficulties when I wanted to figure out 
  which mixtures could  improve <br/> 
  the quality of wine even though the mixtures could be uncorrelatable
  so I decided at the last to make a multivariate plots   <br />
  between the most   correlatble items
  against quality and also choose some uncorrelatble variables and plot
  them against quality to see if there's an intersting trends.
  
* I saw good observations when I plotted alcohol vs chlorides vs quality
  and i was amazed by the results <br/>
  and started to realize from the moment   
  that this industry has many chemical additives that works in a harmony
  together  <br/>
  to derive a better or worse quality
  
  
* In the future I would like to add some attributes like age, place of manufacturing,
  price, taste, smell, fermination process's quality  <br/>
  all these kind of stuff could help me to understand much better the way
  that how chemicals react with to make a good wine.